---
title: "Single-cell RNA-sequencing: variance stabilizing transformations"
author: "Koen Van den Berge"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
---


```{r, echo=FALSE}
if(!"BiocManager" %in% installed.packages()[,1]){
  install.packages("BiocManager")
}
```

# Approximating the variance stabilizing transformation of a Poisson random variable

 - A random variable $Y \sim Poi(\mu)$ has $Var(Y) = E(Y) = \mu$.
 - We are looking for a variance stabilizing transformation (VST) $f(Y)$ such that $Var(f(Y)) = c$, with $c$ any constant. In particular, we need $Var(f(Y))$ to be independent of $\mu$.
 - A first-order Taylor series gives us $f(Y) \approx f(\mu) + (Y - \mu) f'(\mu)$.
 - Rearranging gives us $\left\{ f(Y) - f(\mu) \right\}^2 = (Y-\mu)^2 f'(\mu)^2$.
 - Which may be written as $Var(f(Y)) = Var(Y)f'(\mu)^2$.
 - This shows us that **we want a transformation such that $f'(\mu)^2 = 1/\mu$** because then $\forall \mu: Var(f(Y)) = 1$! Let's find out.
 - The last bullet point can be written as $f'(\mu) = \frac{1}{\sqrt{\mu}}$ and therefore $f(\mu) = \int \frac{1}{\sqrt{\mu}} d\mu = 2 \mu^{1/2}$.
 - Finally, this shows us that the transformation $f(Y) = 2 Y^{1/2}$ ensures $Var(f(Y)) = 1$. Similar, as is often written in the scientific literature, the transformation $f(Y) = Y^{1/2}$ ensures $Var(f(Y)) = 1/4$.
 - Note that these derivations all **rely on the first-order Taylor expansion to be a good approximation**. The VST will therefore work better for random variables with a high mean $\mu$ as the distribution will be more discrete for random variables with a low mean. We show this using simulation below.
 
 
```{r}
set.seed(871)
N <- 1e3
df <- data.frame(mean=rep(NA, N),
                 var=rep(NA, N),
                 transVar=rep(NA, N))
for(kk in 1:N){
  sampledMu <- runif(n=1, min=1, max=500)
  y <- rpois(n=500, lambda=sampledMu)
  df[kk,] <- c(mean(y), var(y), var(sqrt(y)))
}

plot(x=df$mean, y=df$var,
     xlab = "Mean of RV", ylab="Variance of RV")
plot(x=df$mean, y=df$transVar,
     xlab = "Mean of RV", ylab="Variance of transformed RV")

```

**Question**. Why is the first plot heteroscedastic?

<details><summary> Answer. </summary><p>
Remember that $Var(\hat{\mu}) = \frac{\hat{\mu}}{n}$.
Since for all random variables in our simulation $n$ is equal, in our case we have that $Var(\hat{\mu}) = c\hat{\mu}$.
The variance on estimating the mean thus increases with the mean.
</p></details>

# VST for scRNA-seq data based on a negative binomial distribution

## Pearson residuals

As



